---
title: "Introduction to programming with R - Exercises"
output: html_notebook
---


The following code imports sample data points from the carbon project :
```{r}
library(sf)
library(leaflet)
setwd("~/Documents/R base ve R Studio Temel İslemler/organik/")
or <- st_read("Organik_MAdde.shp")
class(or)
or
```



The following code allows us to plot the geographic distribution of data points on an interactive map using leaflet :
```{r}
leaflet(data=or) %>%
  addTiles() %>% 
  addCircles(lng = ~Enlem, lat=~Boylam)

```

## Exercise 1
Extract the names of non-geographic variables in `or` and store it in a variable of type *character*. You have to eliminate the name of the geometry column from the list.
```{r}

```

## Exercise 2
Convert the column *Seviye_3_A* to type *character*.
```{r}

```

## Exercise 3
Create a new *data.frame* object, with only the following 7 columns :
Karbon, OrganikMad, OrganikKar, Kil, Kum, Silt, HacimAgirl. For this exercise, use tools from dplyr package.
```{r}

```

## Exercise 4
Display summary statistics of these 7 columns.
```{r}

```

## Exercise 5
Extract the unique names of the classes stored in column *AraziKulla* and store the information in a character variable.
```{r}

```

## Exercise 6
Calculate the mean values of the same 7 columns separately for column *AraziKulla*. For this, use tools from dplyr package.
```{r}

```

## Exercise 7
Use function *cut()* to divide values of column *Karbon* into 5 classes (5 intervals). Store the resulting factor in data.frame as a new column called *I1*.
```{r}

```

# Exercise 8
Use tools from package *ggplot2* to make a scatterplot values column *OrganikMad* with respect to values of column *Kil*. Use values from column *I1* for dot colours and values from column *Aridite* (eliminate negative Aridite values). 
```{r}

```

# Exercise 9
The map created above has been drawn in geographic coordinates. In this exercise, draw the same map in geographic projections that came with the dataset. 
Hint 1 : Use coordinate columns "POINT_X" "POINT_Y"
Hint 2 : Make sure *leaflet* knows what projection you are using
```{r}

```

# Exercise 10
Modify the map in such a way that the point sizes in the map are proportional to values of column *Silt* and point colors are determined according to class information stored in column *Seviye_1_A*
```{r}

```
