---
title: "Introduction to R"
output: html_notebook
---

# Exercise 1
Download and import the [NOMAD: NASA bio-Optical Marine Algorithm Dataset](https://seabass.gsfc.nasa.gov/wiki/NOMAD). This is a CSV file containing 4458 lines
of measurements of seawater quality collected during several sea campaigns.

- Download the csv file and give the downloaded file the same name
- View contents the first 3 lines of the file 
- Import the file contents into a data.frame called *nomad* using the in-built function *read.table()*
- Print the names of the columns of the newly created data.frame
- Use the first 6 columns of the data.frame to create a new column of type *POSIXct*. Hint : Generate a character variable of date/time in %Y/%m/%d %H:%M:%S" format using *paste()*. Then convert the string to *strptime()* and call it datetime1
- Make sure that *datetime* variable is of class *POSIXct*
- Remove the first 6 columns of *nomad*
- Prepend *datetime* as a new column to *nomad*. Call new column as "datetime"
- Convert all invalid numerical measurements to *NA*
- Make a scatterplot with function *plot()* where the column *chl_a* is plotted with respect to column *lw489*. Use red dots (instead of black hollow circles) of size 3. Hint : Consult documentation of function *par()*.


```{r}
urlNomad = "https://raw.githubusercontent.com/scizmeli/Red/master/nomad_seabass_v2.a_2008200.txt"
```

#Download the csv file
```{r}
#
```
download.file(urlNomad, "nomad_seabass_v2.a_2008200.txt")

#Import data.frame
```{r}
#
```
nomad <- read.table("nomad_seabass_v2.a_2008200.txt",header=T,sep=",", comment.char = "!")

#Make a datetime variable of tyep Posixct
date1 = paste(nomad$year, nomad$month, nomad$day,sep="/")
time1 = paste(nomad$hour, nomad$minute, nomad$second,sep=":")
datetime1 = paste(date1, time1)
datetime = strptime(datetime1,"%Y/%m/%d %H:%M:%S")
datetime = as.POSIXct(datetime)
class(datetime)
nomad = nomad[,-(1:6)] 
nomad = cbind(data.frame(datetime=))
nomad[nomad==-999]=NA
plot(nomad$lw489, nomad$chl_a, pch=".",col="red",cex=3)
```{r}
#
```

# Remove the first 6 columns of the data frame
```{r} *chl_a* is plotted with respect to column *lw489*
#
```

# Prepend *datetime* as a new column to *nomad*. Call new column as "datetime"
```{r}
#
```

# Convert all invalid numerical measurements to *NA*
```{r}
#
```

# Make a scatterplot of *chl_a* with respect to *lw489*
```{r}
#
```

